LHC abort gap monitor electronics upgrade

### 2024-05-11 paper.tex
Update on Overleaf.

\section{Introduction}
During its operation at flat top energy, the LHC ring stores up to \SI{350}{MJ} of beam energy, spread across 2808 bunches with an average beam transverse size under \SI{1}{mm}.
%
This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}.
%
To safely extract the beam from the circulating beam lines, a series of kicker magnets belonging to the beam dump system is used. 
%
Kicker magnets need \SI{3}{\micro\second} to ramp up their magnetic field~\cite{beam_dump_system}.
%
This defines the size of an Abort Gap (AG) in the bunch train.
%
Due to system imperfections, a residual amount of particles populates this gap.
%
The typical AG population levels are two to three orders of magnitude lower in comparison to a single pilot LHC bunch.
%
When the beam dump is initiated, the particles in the AG do not receive full
extraction force, are lost on their way to the dump line and can generate damages or magnet quenches.  
%
%That causes the particles in the AG to
%are deflects towards the vacuum chamber, possibly causing the magnets to quench. 
%
The purpose of the AGM \cite{particles_in_ag} is to 
a~continuous monitoring of the AG population in order to trigger the abort gap cleaning when  necessary. 
\\
This paper discusses the upgrade of the operational AGM installed in the LHC since 2010, detailing the system architecture and analog electronics chain, and provide a first performance comparison  of both systems.

\section{System architecture}
The system architecture is shown in Fig.~\ref{fig:bsra_chain}.
%
It depicts the layout of the LHC, with beam 1 represented in blue and beam 2 in red.
%
The LHC AGM consists of three main parts:
%
\begin{enumerate}
    \item \textbf{an optical front-end (OF)}, which includes~a pre-amplifier, located in the LHC tunnel,
    \item \textbf{a data acquisition system (DAQ)} deployed on an FPGA platform located outside the radiation area,
    \item \textbf{a control software (SW)} that retrieves, interprets, and publishes the measured data.
\end{enumerate}
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{Overview of an abort gap monitor system.}
    \label{fig:bsra_chain}
\end{figure}

@@

%The particle population in the AG is measured intercepting the synchrotron radiation (SR). 
%
The AGM is designed to collect part of th extracted SR on  a Multi-channel photo-multiplier (PMT), which converts it  into an electric signal.
%
Neutral Density filters in the optical front-end (OF) allow  changing the light intensity intercepted by the PMT, thus optimize the system dynamic range.
% 
Subsequently, the electric signal is amplified, and transmitted over a \SI{100}{m}-long cable to a Data Acquisition (DAQ) system.
%
The DAQ integrates the incoming signal in \SI{100}{\nano\second} intervals within a \SI{6.4}{\micro\second}-long acquisition window.
%
Each of the sixty-four \SI{100}{\nano\second} integrals is independently averaged in an Exponential Moving Averager (EMA).
%
The averaged result is published to the software with a repetition rate of 10~Hz, which is sufficient to react to even the fastest changes in AG population.~\cite{high_sensitivity_measurement}.
%
\\
The measurement period is determined by the PMT used.
%
In our case, it is HAMAMATSU R5916U-50. 
%
Its maximum duty cycle is \SI{1}{\%}. 
%
With a gate pulse duration of \SI{3}{\micro\second} and one LHC revolution period of \SI{89}{\micro\second}, the measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}).
%
This corresponds to a duty cycle of \SI{0.84}{\%}.
\\
%

@@

The currently used system is shown in Fig.~\ref{fig:current_system_processing}.
%
The optical path, including filters, is common to both the operational version of the system, and the proposed upgrade. As well, the PMT type is the same in both systems.

The operational setup uses a digital acquisition board (DAB64x) equipped by Altera Stratix EP1S40 FPGA, used for the data processing.
%
The DAB64x hosts two Individual Bunch Measurement Systems (IBMS)~\cite{ibms}, integrating the analogue input signal via the LHCb2002 ASIC~\cite{lhcb_asic}.
%
The analogue signal is provided by MCP PMT, and it is amplified before the integration to improve on the measurement SNR.
%
The amplifier used is HAMAMATSU C5594~\cite{bsra_first_operation}, featuring a fixed voltage gain of \SI{36}{dBV}, in the bandwidth of \SI{50}{kHz} to \SI{1.5}{GHz}.
%
The LHCb2002 in the IBMS integrates the input amplified signal in \SI{25}{\nano\second} intervals, producing a stream of analogue 'DC-like' values.
%
These are subsequently sampled by a 40MSPS 14-bit ADC, resulting in a digital data stream, where each value corresponds to an integral of an individual LHC bunch.
%
The DAQ gateware sums four consecutive samples to compute the
bunch charge present on a~\SI{100}{ns} interval.
%
This process generates 64 values per single turn acquisition.
%
For each individual value, a standard deviation and mean value is calculated, and separately filtered by EMA. 
%
All collected data are stored in a RAM block shared to a front-end computer (FEC) via VME64 bridge.
%
The front-end acts as a server, publishing the data upstream through FESA~\cite{fesa}.

\subsection{Reasons for Upgrade}
The present AGM uses old technology to measure the beam charge. 
%
The analogue integrators used in IBMS are manufactured using obsoleted technology, rendering them impossible to get.
%
In addition, each integrator is unique: different gain, different switching characteristics, different offsets. 
%
So, the gateware must correct these effects.
%
Further, the amplifier bandwidth (50 kHz-1,5 GHz) increases system complexity, as a correction for the missing DC component needs to be implemented.
%
The amplifier gain is also fixed, so the ADC cannot optimally re-use available dynamic range.
\\
The system relies on an outdated FPGA platform, and the development SW is no more compatible with the latest operating systems.

\section{System Upgrade}
To address the long-term maintenance concerns, the original system described above has to be refurbished. 
%
While the optics of the new system was copied from the operational one, the DAQ and the front-end amplifier were replaced by a modern solution.
%
A new DC-coupled amplifier was designed, allowing for dynamic change of the gain.
%
Original analogue integrators were replaced by a 4-channel 500MSPS 14-bit FMC-form factor sampler.
%
FMC ADC module is installed into an in-house built FPGA carrier (VFC-HD), featuring 2$\times 8$GBits DDR3 memories and a large Arria V FPGA.
%
Such solution eliminates completely the need for analogue integrators, as all the signal processing can be implemented in the gateware.  
\\
An updated version of the AGM was in 2023 installed \textit{in parallel to the actually running AGM}, so that we can compare behavior of both systems.

@@

Figure \ref{fig:amp_schematics} depicts the block diagram of the new amplifier.
%
The amplifier was designed to match the used PMT, while being highly configurable.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
%
A~UART interface allows for setting a transimpedance gain.
%
The gain is selectable in a range of \SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}.
%
This corresponds to a~voltage gain of \SI{41}{dB} to \SI{87}{dB} in a $50\Omega$ system.
%
A programmatic DC-offset compensation is implemented in the amplifier.
%
It can be used - together with variable gain - to optimize the signal's amplitude for the ADC.
%
The amplifier implements an alarm, which interlocks the PMT gate when the incident light generates a current, that is near the PMT's maximum allowable average current.
%
The alarm threshold setting is stored through UART into the amplifier's non-volatile memory, so that it protects the PMT immediately at the amplifier power-up.
%
The alarm status is also transmitted to the gateware.
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}
%
\subsection{Acquisition system}
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Acquisition system schematics.}
    \label{fig:bsra_daq}
\end{figure}
%
The new DAQ (Fig.~\ref{fig:bsra_daq}) receives asynchronous data samples as a stream of four frames every~\SI{8}{\nano s}, each corresponding to one channel.
%
Each frame contains 4$\times$16-bit words, storing left-aligned 14-bit ADC samples. 
%
To compute integrals over 4$\times$~\SI{25}{\nano\second} bunch slots, the system employs data tagging. 
%
The tagger resamples LHC timing, and generates timing masks, allowing for identification of bunch and turn starts within each individual ADC frame.
%
Tags are written into 2 unused LSBs of each word to indicate the position of the start of the bunch or turn.
\\
%
The tagged samples are stored into a~FIFO, and integrated. 
%
The integrator begins summing values upon encountering the first bunch tag appearing within the acquisition window.
%
In parallel, it reads a stream of 64~values of a stored 'offset' from a pedestal memory (P[63:0]).
%
The offset is on-fly subtracted from the integrated data stream. 
%
Such pedestal subtraction is needed to remove systematic effects, e.g., PMT gate opening. 
%
The output values are filtered by an EMA, results are kept in RAM, and read out through the VME64x bus~\cite{my_thesis}.

@@

The new installation is depicted \textit{in retracted position} in Fig.~\ref{fig:installation}, highlighted by a~violet frame.
%
Adjacent to its right is the operational AGM, while the longitudinal density monitor (LDM) occupies the left side.
%
All instruments share the same light source provided by the LHC beam synchronous light telescope (BSRT).
%
The red arrow indicates the direction of the light entering the optical table.
%
The light is unequally split using multiple reflective foils into their respective destinations such, that the new AGM intercepts only 10\% of luminosity of the operational AGM.
%
Both AGMs are equipped with LEDs installed atop of their reflective foils.
%
These serve to perform functional tests.
%
The new AGM amplifier is mounted on the top of the optical front-end to ease the interventions.

@@

\vspace{-5mm}
The initial measurements were conducted in November 2023 using the LHC ion beam.
%
Figure~\ref{fig:comparison_chart} shows the very first measurements, compared to the operational AGM.
%
Both measurements agree to 2\% within the operational area, signals however deviate starting at $47^{\text{th}}$ bin, resulting in maximum relative error of 4\%.
%
This is surprising, as no signal shall be present - the PMT gates are closed. 
%
However, the operational AGM signal gradually decays, and extends below zero.
%
This effect is systematic, and is mostly caused by analogue signal processing chain and non-ideal analogue integration.
%
The new system effectively eliminates this.

>>>>

\section{Introduction}
The LHC ring stores up to \SI{350}{MJ} of beam energy, when operating at the flat top energy, spread across 2808 bunches with an average beam transverse size under \SI{1}{mm}.
%
This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}.
%
To safely extract the beam from the circulating beam lines, a series of kicker magnets belonging to the beam dump system is used. 
%
Kicker magnets need \SI{3}{\micro\second} to ramp up their magnetic field~\cite{beam_dump_system}.
%
This defines the size of an Abort Gap (AG) in the bunch train.
%
Due to system imperfections, a residual amount of particles populates this gap.
%
The typical AG population levels are two to three orders of magnitude lower in comparison to a single pilot LHC bunch.
%
When the beam dump is initiated, the particles in the AG do not receive full
extraction force, leading to losses along the dump line that could result in damage or magnet quenching.
%
%That causes the particles in the AG to
% deflect towards the vacuum chamber, possibly causing the magnets to quench. 
%
The purpose of the AGM is a~continuous monitoring of the AG population in order to trigger the abort gap cleaning when  necessary \cite{particles_in_ag} .
\\
This paper discusses the upgrade of the operational AGM installed in the LHC since 2010, detailing the system architecture and analogue electronics chain, and provides a first performance comparison  of both systems.

\section{System architecture}
The system architecture is shown in Fig.~\ref{fig:bsra_chain}.
%
It depicts the layout of the LHC, with beam 1 represented in blue and beam 2 in red.
%
The LHC AGM consists of three main parts:
%
\begin{enumerate}
    \item \textbf{optical front-end (OF)}, which includes~a pre-amplifier, located in the LHC tunnel,
    \item \textbf{data acquisition system (DAQ)} deployed on an FPGA platform located outside the radiation area,
    \item \textbf{control software (SW)} that retrieves, interprets, and publishes the measured data.
\end{enumerate}
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{Overview of an abort gap monitor system.}
    \label{fig:bsra_chain}
\end{figure}

@@

%The particle population in the AG is measured intercepting the synchrotron radiation (SR). 
%
The AGM collects part of the extracted SR via Multi-channel plate photo-multiplier (MCP-PMT)converting it into an electric signal.
%
Neutral density filters in the optical front-end (OF) allow  changing the light intensity intercepted by the MCP-PMT, thus optimizing the system dynamic range.
% 
Subsequently, the electric signal is amplified and transmitted over a \SI{100}{m}-long cable to DAQ.
%
The DAQ integrates the incoming signal in \SI{100}{\nano\second} intervals within a \SI{6.4}{\micro\second}-long acquisition window.
%
Each of the 64 \SI{100}{\nano\second} integrals is independently averaged in an Exponential Moving Averager (EMA).
%
The averaged result is published to the software with a repetition rate of \SI{10}{Hz}, which is sufficient to react to even the fastest changes in AG population.~\cite{high_sensitivity_measurement}.
%
\\
The measurement period depends on the specific MCP-PMT (HAMAMATSU R5916U-50) that features a maximum duty cycle of \SI{1}{\%}. 
%is determined by the PMT used.
%
%In our case, it is HAMAMATSU R5916U-50. 
%
%Its maximum duty cycle is \SI{1}{\%}. 
%
With a gate pulse duration of \SI{3}{\micro\second} and one LHC revolution period of \SI{89}{\micro\second}, the measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}).
%
This corresponds to a duty cycle of \SI{0.84}{\%}.
\\
%

@@

The currently used system is shown in Fig.~\ref{fig:current_system_processing}.
%
The optical path, including filters and MCP-PMT, is common to both the operational version of the system, and the proposed upgrade.
%As well, the PMT type is the same in both systems.
The operational setup utilizes a digital acquisition board (DAB64x) incorporating an Altera Stratix EP1S40 FPGA for data processing.
%
The DAB64x hosts two Individual Bunch Measurement Systems (IBMS)~\cite{ibms}, integrating the analog input signal via the LHCb2002 ASIC~\cite{lhcb_asic}.
%
The analog signal generated by  the MCP-PMT is amplified  prior to integration to  enhance the measurement's Signal-to-Noise Ratio  (SNR).
%
The amplifier (HAMAMATSU C5594~\cite{bsra_first_operation}) has a fixed voltage gain of \SI{36}{dBV} and the bandwidth of \SI{1.5}{GHz} with the low cutoff frequency at \SI{50}{kHz}.
%
The amplified signal is fed to the LHCb2002 in the IBMS integrating in \SI{25}{\nano\second} intervals, producing a stream of analog  'DC-like' values.
%
These are subsequently sampled by a \SI{40}{MSPS} 14-bit ADC, resulting in a digital data stream, where each value corresponds to an integral of an individual LHC \SI{25}{ns} bucket.
%
The DAQ gateware sums four consecutive samples to compute the
proton charge present on a~\SI{100}{ns} interval.
%
This process generates 64 values per single turn acquisition.
%
For each individual value, a standard deviation and mean value is calculated, and separately filtered by EMA. 
%
All collected data are stored in a RAM block shared to a front-end computer (FEC) via VME64 bridge.
%
A dedicated  front-end  server, publishes the data stream through FESA~\cite{fesa}.

\subsection{Reasons for Upgrade}

The operational AGM primarily relies on outdated   technology, such as  the analogue integrators used in the IBMS,  which are no longer manufactured.
%that are not produced anymore.  manufactured using obsoleted technology, rendering them impossible to get.
%
Furthermore, each integrator is unique, with varying gain, switching characteristics, and offsets. The gateware must correct these variations.
%Furthermore,  each integrator is unique, with varying gain, switching characteristics, and offsets. 
%
%The gateware must correct these variations.
%different gain, different switching characteristics, different offsets and the 
%
%So, the
%gateware must correct these effects.
%
In addition, the amplifier bandwidth (50 kHz-1,5 GHz) %increases system complexity, as
required a correction for the missing DC component and the amplifier gain is fixed,  for which  the overall dynamic range is limtied by the ADC. 
%needs to be implemented.
%
%dynamic range is not fully exploited. cannot optimally re-use available dynamic range.
\\
The system relies on an outdated FPGA platform, and the development SW is no more compatible with the latest operating systems.

\section{System Upgrade}
To address the long-term maintenance concerns, the original system described above has to be refurbished. 
%
While the optics of the new system was copied from the operational one, the DAQ and the front-end amplifier were replaced by a modern solution.
%
A new DC-coupled amplifier was designed, thus allowing enahncing the overall dynamic range. 
%
It was decided to change from analog to digital instegration of the PMT signal, via a DAQ based on  a 4-channel \SI{500}{MSPS} 14-bit FMC-form factor sampler.
%
The FMC ADC module is installed into an in-house built FPGA carrier (VFC-HD), featuring 2$\times 8$GBits DDR3 memories and a large Arria V FPGA.
%
Such a solution eliminates completely the need for the obsolete analogue integrators, as all the signal processing can be implemented in the gateware.  
\\
In early 2023, the new AGM system was installed alongside the legacy system to facilitate an easy comparison of both systems' performance.
%An updated version of the AGM was in 2023 installed \textit{in parallel to the actually running AGM}, so that we can compare behavior of both systems.

@@

Figure \ref{fig:amp_schematics} depicts the block diagram of the new amplifier.
%
The amplifier was designed to match the AGM MCP-PMT and to be  highly configurable.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
%
A universal asynchronous receiver-transmitter (UART) interface allows for setting a trans-impedance gain.
%
The gain is adjustable in a range of \SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}.
%
This corresponds to a~voltage gain of \SI{41}{dB} to \SI{87}{dB} in a $50\Omega$ system.
%
A programmable DC-offset compensation is implemented in the amplifier.
%
It can be used - together with variable gain - to optimize the signal's amplitude for the ADC.
%
The amplifier design incorporates an alarm. This alarm interlocks the MCP-PMT gate when the incident light produces a current nearing the MCP-PMT's maximum allowable average current.
%The amplifier implements 
%an alarm that interlocks the MCP-PMT gate when the incident light generates a current approachng  the MCP-PMT's maximum allowable average current.
%
The alarm threshold setting is stored through the UART into the amplifier's non-volatile memory, so that 
%it protects 
the MCP-PMT is protected also after a power reset. %immediately at the amplifier power-up.
%
The alarm status is also transmitted to the gateware.
%
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier.}
    \label{fig:amp_schematics}
\end{figure}
%
\subsection{Acquisition system}
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Acquisition system schematics.}
    \label{fig:bsra_daq}
\end{figure}
%
The new DAQ (Fig.~\ref{fig:bsra_daq}) receives asynchronous data samples as a stream of four frames every~\SI{8}{\nano s}, each corresponding to one channel.
%
Each frame contains 4$\times$16-bit words, storing left-aligned 14-bit ADC samples. 
%
To compute integrals over 4$\times$~\SI{25}{\nano\second} bunch slots, the system employs data tagging. 
%
The tagger resamples the LHC timing, and generates timing masks that  provide each individual ADC frame with %allowing for identification of 
bunch and turn start markers. 
%within each individual ADC frame.
%
Tags are inserted into the  2 unused least significan bits (LSBs) of each word to indicate the start of the  the bunch or turn.
\\
%
The tagged samples are stored in a~FIFO ((First In, First Out), and then integrated. 
%
The integrator is triggered %begins summing values 
by the  the first bunch tag appearing within the acquisition window.
%
In parallel to the integral (sum of the values), the system reads a stream of 64~values of a stored 'offset' from a pedestal memory (P[63:0]).
%
The offset is subtracted  from the integrated data stream on the fly.
%
Such pedestal subtraction is needed to remove systematic effects, e.g. related to the  MCP-PMT gate opening. 
%
The output values are filtered by an Exponential Moving Average (EMA), the results are stored  in RAM, and then read out through the VME64x bus~\cite{my_thesis}.

@@

The new installation, when \textit{in retracted position} is  highlighted by a~violet frame in Fig.~\ref{fig:installation}, 
%
The legacy AGM  (on the right) and  the longitudinal density monitor (LDM, on the left) are also partially visible in the same figure. % Adjacent to its right is the operational AGM, while the longitudinal density monitor (LDM) occupies the left side.
%
All these instruments are part of the LHC beam synchronous light telescope (BSRT) and use part of the synchrotron radiation extracted by the in-vacuum extraction system sitting above the telescope. %  instruments share the same light source provided by the LHC beam synchronous light telescope (BSRT).
%
The red arrow indicates the direction of the light entering the optical table.
%
The light is unequally split using multiple reflective foils into their respective destinations.  The new AGM receives  only 10\% of the radiation sent to the legacy AGM.
%
Both AGMs are equipped with LEDs installed  on top of their reflective foils for functional and calibration tests. 
%
%These serve to perform functional tests.
%
It is worth noting  that the new AGM amplifier is mounted on top of the optical front-end to facilitate  interventions.

@@

\vspace{-5mm}
After few tests and few adjustements in the system layout The initial measurements were conducted in November 2023 using the LHC ion beam.
%
Figure~\ref{fig:comparison_chart} shows the very first measurements, compared to the operational AGM.
%
Both measurements agree to 2\% within the operational area, signals however deviate starting at $47^{\text{th}}$ bin, resulting in maximum relative error of 4\%.
%
This is surprising, as no signal shall be present - the MCP-PMT gates are closed. 
%
However, the operational AGM signal gradually decays, and extends below zero.
%
This effect is systematic, and is mostly caused by analog signal processing chain and non-ideal analogue integration.
%
The new system effectively eliminates this.
